Realizing temporal graphs from fastest travel times

### 2023-09-21 STACS2023/nina_shortversion.tex
Shorter version in progress

In a graph $G=(V,E)$, a \emph{feedback edge set} $F \subset E(G)$ is a subset of edges, such that each cycle in $G$ has at least one edge in $F$.
The minimum such set $F$ is called a  \emph{minimum feedback edge set} and its size, $k = |F|$, is called the \emph{feedback edge number} of graph $G$.
Note that one can find a minimum feedback edge set in linear time, by calculating a spanning tree (or forest) $T$ of the given graph $G$ and then removing all of the edges $T$ from $G$.%, \ie $F = E(G) \setminus E(T)$.

@@

We first give an informal and very high-level description of the main ideas of the algorithm.
We then present the first part in more details and deffer the reader to the appendix, for the remaining part and correctness proof.
It is well-known that the number of paths between two vertices in a static graph can be upper-bounded by a function of the feedback edge number of the graph. Intuitively, this would allow us to ``guess'' paths between a small number of vertex pairs in $G$ (the underlying graph of our instance \deltaExact). These paths would then be used to create (representatives of) fastest temporal paths among vertex pairs in $(G,\lambda)$. We would then try to use the polynomial algorithm for trees (which includes paths) to figure out the labeling $\lambda$ of the edges used by these path. An idea like this has been used for several other temporal graph problems related to the connectivity between two vertices~\cite{casteigts2021finding,FMNR22a,EMM22}.

In our case, however, it is not enough to consider fastest temporal paths among some subset of vertices, but we have to determine the fastest temporal paths between all pairs of vertices. This means that we have to somehow extend the labeling $\lambda$ of the edges used by guessed paths to all other edges of $G$.
Besides that we also need to make sure that all temporal paths that do not follow our guessed paths are not too fast.

	A further difficulty arises from the somewhat unintuitive characteristic of fastest paths in temporal graphs.
 When considering the problem of finding shortest paths among vertices in static graphs, we have the nice property that if a shortest path $P$ from vertex $u$ to $v$ traverses the vertex $w$,
	then the subpath of $P$ from $u$ to $w$ is a shortest path from $u$ to $w$. This can help tremendously when developing algorithms.
	However, one can see that this property does not carry over to fastest paths in temporal graphs, for an example see~\cref{fig:ftpExample}.
 For us this means that even if we guess a fastest path between two vertices, we cannot simply determine all labels of the edges used by the fastest path.
	%This makes all the standard intuition and work with these kind of paths, much harder.
    %This also explains why our FPT algorithm, w.r.t the feedback edge number, is so complex. 
    %Usually, the feedback edge number parameter (denote it $k$) provides an easy algorithm, 
    %where one first determines $O(k)$ special vertices,
    %guesses desired structures among them (\ie iterates through all possibilities) and then simply extends the solution to the rest of the graph.
    %An example with this approach, on temporal graphs, can be observed in \cite{casteigts2021finding} where the authors provide an FPT algorithm that determines all temporal paths, under certain constraints.
    %In our case the similar approach cannot work, as we cannot at any point calculate some partial labeling $\lambda$ of a delta periodic graph $(G,\lambda)$,
    %that could at the end be extended to the whole $G$.

@@

To overcome the above mentioned difficulties, we propose the following high-level strategy.
\begin{itemize}
    \item We identify a small number of ``interesting vertices'' between which we guess representative fastest paths.
    \item We show that we can determine most labels along each path up to an additive constant. We show that the number of labels we cannot determine along such a path is small. 
    \item For each additive constant and each non-determined label along such a path we introduce a variable for an Integer Linear Program (ILP).
    \item We use constraints in the ILP to ensure that alternative connections (not using the guessed fastest path) between vertex pairs are not too fast.
    \item We compute a solution to each of the ILPs (if it exists) and construct a labeling from that solution. Since the ILP has few variables and not too many constraints, this computation can be done efficiently using Lenstra's Theorem~\cite{Lenstra1983Integer}.
\end{itemize}
It is important to point out that each guess of fastest paths, gives rise to one ILP. Since there are many guesses ($f(k)$ many, where $f$ is a double exponential function), we get many ILPs.
If none of the ILP instances gives a positive solution, then there exists no labeling realizing given matrix.
This is true because considering all guesses of fastest temporal paths is equivalent to exhaustively searching through all possible temporal paths between vertices.

@@

%Let us remember the following basic property of the problem \deltaExact.
Recall that a labeling $\lambda$ of $G$ satisfies $D$ if the duration of a fastest temporal path from vertex $v_i$ to $v_j$ equals $D_{v_i, v_j}$.
%, and no other temporal path from $v_i$ to $v_j$ is faster.
In order to find a labeling that satisfies this property we split our analysis in nine cases.
We consider fastest temporal paths 
where $v_i$ is in one of the sets $U,V(G') \setminus U, Z$,
and $v_j$ is in one of the sets $U,V(G') \setminus U, Z$.
In each of these cases we guess the underlying path $P$ that at least one fastest temporal path from $v_i$ to $v_j$ follows, 
which results in one equality constraint for the labels on the path $P$. 
For all other temporal paths from $v_i$ to $v_j$ we know that they cannot be faster, so we introduce inequality constraints for them.
This results in producing $poly(n,m)$ constraints. 
Note that we have to do this while keeping the total number of variables upper-bounded by some function in $k$.

For an easier understanding and the analysis of the algorithm we give the following definition.
\begin{definition}\label{def:segments}
	Let $U \subseteq V(G')$ be a set of vertices of interest and let $u,v \in U$.
	A path $P = (u=v_1,v_2, \dots, v_p = v)$  in graph $G'$, where all inner vertices are not in $U$, \ie $v_i \notin U$ for all $i \in \{ 2, 3, \dots, p-1\}$,
	is called a \emph{segment} from $u$ to $v$. We denote it as $S_{u,v}$.
\end{definition}
%From the definition it follows, that there are $O(k^2)$ segments in $G'$. \todo{True but does not follow directly just from the definition}
Note from \cref{def:segments} we get that $S_{u,v} \neq S_{v,u}$, since we consider paths to be directed. Observe that a temporal path in $G'$ between two vertices of interest is either a segment, or consists of a sequence of some segments. Furthermore, since we have at most $4k$ interesting vertices in $G'$, we can deduce the following important result.
\begin{corollary}\label{obs:FPT-k2segments}
	There are at most $O(k^2)$ segments in $G'$.
\end{corollary}
%In the analysis of the algorithm we rely heavily on the fact that there are $O(k^2)$ segments in $G'$.
%, which we prove later on (see~\cref{obs:FPT-k2segments}).
%For a pair $u,v$ of vertices in $U$, there are $O(k!)=O(k^k)$ possible paths in $G'$ between them.

@@

	Let $S_{u,v} = (u=v_1,v_2, \dots, v_p = v)$ and $S_{w,z} = (w=z_1,z_2, \dots, z_r = z)$ be two segments,
	first one between vertices $u,v \in U$, second one between vertices $w, z \in U$.

@@

	Recall that $S_{u,v}\neq S_{v,u}$ for every $u,v\in U$. Furthermore note that we did not assume that the sets $\{u,v\}$ and $\{w,z\}$ are disjoint. Therefore, by repeatedly making the above guesses, we also guess the following fastest temporal paths: 

@@

	and let $w \in U \cup Z^*$ be either a vertex of interest or a representative vertex of a tree, whose clipped vertex is a vertex of interest.

@@

	For a fixed segment $S_{u,v}$ and a fixed vertex $w \in U \cup Z^*$ we have $O(k^k)$ different possible such paths, therefore we make $O(k^{k^4})$ guesses for these paths.

@@

	In the case when $S_{u,v}$ is of length $4$, the fixed vertex $v_i$ is the middle vertex (\ie vertex that is on distance two from $u$ and $v$ in $S_{u,v}$).
	If $S_{u,v}$ is not of length~$4$, we fix an arbitrary vertex $v_i \in S_{u,v} \setminus \{u,v\}$.

@@

	Note that all inner vertices of $S_{w,z}$ (\ie vertices in $S_{w,z} \setminus \{w,z\}$)
	can be reached from $v_i$ by some path that has to go through $w$ or $z$.
	Therefore, the values of the duration of the fastest path from $v_i$ to vertices on 
	the segment
	$S_{w,z}$ have to increase up to a certain point, when traversing the segment coming through $w$ and coming through $z$.

@@

		Note that in this case we do not introduce any new guesses, as we have aready guessed 

@@

	Let $w \in U \cup Z^*$ be either a vertex of interest or a representative vertex of a tree, whose clipped vertex is a vertex of interest,
	and let $S_{u,v} = (u=v_1,v_2, \dots, v_p = v)$ be a line segment in $G'$.

@@

	All together we make two guesses for each pair of vertex $w \in U$ and segment $S_{u,v}$.
	We repeat this for all vertices of interest, and all segments,
	which produces $O(k^{k^2})$ new guesses.
	%
	For an example see~\cref{fig:FPT-guessG7}.
	Detailed analysis follows arguing from above (as in G-\ref{FPT:guess-splitFromAnotherSegmentAndPaths}) and is deferred to Appendix.
	% counter for guesses continues
	\setcounter{guesscounter}{\value{enumi}}
\end{enumerate}
%
\begin{figure}[t]
	\centering
	\begin{subfigure}[b]{0.48\textwidth}
		\centering
		\resizebox{\linewidth}{!}{
			\includegraphics{fig-guess1}}
		\caption{Example of a guess G-\ref{FPT-guessFTPamongv2z2}, where we guessed the fastest temporal paths of the form $v_2 \rightarrow u \leadsto w \rightarrow z_2$ (in blue)
			and $v_2 \rightarrow u \leadsto z \rightarrow z_{r-1}$ (in red).
			\label{fig:FPT-guessG4}}
	\end{subfigure}
	\quad
	\begin{subfigure}[b]{0.48\textwidth}
		\centering
		\resizebox{\linewidth}{!}{

			\includegraphics{fig-guess2}}
		\caption{Example of a guess G-\ref{FPT:guess-uToSegmentz2}, where we guessed the fastest temporal paths of the form $w \leadsto u \rightarrow v_2$ (in blue) and $w \leadsto v \rightarrow v_{p-1}$ (in red). 
			\label{fig:FPT-guessG5}}
	\end{subfigure}

	\begin{subfigure}[b]{0.48\textwidth}
		\centering
		\resizebox{1.2\linewidth}{!}{
			%
			\includegraphics{fig-guess3}}
		\caption{Example of a guess G-\ref{FPT:guess-splitFromAnotherSegmentAndPaths}, where, for a fixed vertex $v_i \in S_{u,v}$,
			we calculated its corresponding split vertex $z_j \in S_{w,z}$,
			and guessed the fastest paths of the form
			$v_i \rightarrow v_{i-1} \rightarrow \cdots \rightarrow u \leadsto z \rightarrow z_{r-1} \cdots \rightarrow z_j$ (in blue) 
			and $v_i \rightarrow v_{i+1} \rightarrow \cdots \rightarrow v \leadsto w \rightarrow z_2 \rightarrow \cdots \rightarrow z_{j-1}$ (in red). 
			\label{fig:FPT-guessG6}}
	\end{subfigure}
	\quad
	\begin{subfigure}[b]{0.48\textwidth}
		\centering
		\resizebox{\linewidth}{!}{
			\includegraphics{fig-guess4}}
		\caption{Example of a guess G-\ref{FPT:guess-splitFromUtoAnotherSegment}, where, for a vertex of interest $w$, 
			we
			calculated its corresponding split vertex $v_i \in S_{u,v}$,
			and guessed the fastest paths of the form
			$w \leadsto u \rightarrow v_2 \rightarrow \cdots \rightarrow v_i$  (in blue) 
			and $w \leadsto v \rightarrow v_{p-1} \rightarrow \cdots \rightarrow v_{i+1}$ (in red). 
			\label{fig:FPT-guessG7}}
	\end{subfigure}
	\caption{An example of guesses G-\ref{FPT-guessFTPamongv2z2}, G-\ref{FPT:guess-uToSegmentz2}, G-\ref{FPT:guess-splitFromAnotherSegmentAndPaths} and G-\ref{FPT:guess-splitFromUtoAnotherSegment}.}
\end{figure}
%
We create all of the above guesses independently.
We start with the first one, that results in $O(k^{k^3})$ different possibilities, then we split each one of these guesses into $O(k^{k^3})$ new ones, that respond to the guessing in the second step, etc.
After creating all of the guesses we end up with $f(k)$ different cases (where $f$ is a double exponential) 
and create one ILP instance for each case.

>>>>

In a graph $G=(V,E)$, a \emph{feedback edge set} $F \subset E(G)$ is a subset of edges, such that each cycle in $G$ has at least one edge in $F$.
The minimum such set $F$ is called a  \emph{minimum feedback edge set} and its size, $k = |F|$, is called the \emph{feedback edge number} of graph $G$.
Determining $F$ can be done in linear time, by simply removing edges of a spanning tree (forest) from the given graph.
%one can find a minimum feedback edge set in linear time, by calculating a spanning tree (or forest) $T$ of the given graph $G$ and then removing all of the edges $T$ from $G$.%, \ie $F = E(G) \setminus E(T)$.

@@

We first give an informal and very high-level description of the main ideas of the algorithm.
We then present the first part in more details and deffer the reader to the appendix for the remaining part and correctness proof.
It is well-known that the number of paths between two vertices in a static graph can be upper-bounded by a function of the feedback edge number of the graph. Intuitively, this would allow us to ``guess'' paths between a small number of vertex pairs in $G$ (the underlying graph of our instance \deltaExact). These paths would then be used to create (representatives of) fastest temporal paths among vertex pairs in $(G,\lambda)$. We would then try to use the polynomial algorithm for trees (which includes paths) to figure out the labeling $\lambda$ of the edges used by these path. An idea like this has been used for several other temporal graph problems related to the connectivity between two vertices~\cite{casteigts2021finding,FMNR22a,EMM22}.
%
%In our case, however, it is not enough to consider fastest temporal paths among some subset of vertices, but we have to determine the fastest temporal paths between all pairs of vertices. This means that we have to somehow extend the labeling $\lambda$ of the edges used by guessed paths to all other edges of $G$. Besides that we also need to make sure that all temporal paths that do not follow our guessed paths are not too fast.

In our case, however, this approach does not work. Mainly because of the
	%A further difficulty arises from the 
 somewhat unintuitive characteristic of fastest paths in temporal graphs.
 When considering the problem of finding shortest paths among vertices in static graphs, we have the nice property that if a shortest path $P$ from vertex $u$ to $v$ traverses the vertex $w$,
	then the subpath of $P$ from $u$ to $w$ is a shortest path from $u$ to $w$. This can help tremendously when developing algorithms.
	However, one can see that this property does not carry over to fastest paths in temporal graphs, for an example see~\cref{fig:ftpExample}.
 For us this means that even if we guess a fastest path between two vertices, we cannot simply determine all labels of the edges used by the fastest path.
 Besides that, it is not enough to consider fastest temporal paths among some subset of vertices, but we have to determine the fastest temporal paths between all pairs of vertices.
	%This makes all the standard intuition and work with these kind of paths, much harder.
    %This also explains why our FPT algorithm, w.r.t the feedback edge number, is so complex. 
    %Usually, the feedback edge number parameter (denote it $k$) provides an easy algorithm, 
    %where one first determines $O(k)$ special vertices,
    %guesses desired structures among them (\ie iterates through all possibilities) and then simply extends the solution to the rest of the graph.
    %An example with this approach, on temporal graphs, can be observed in \cite{casteigts2021finding} where the authors provide an FPT algorithm that determines all temporal paths, under certain constraints.
    %In our case the similar approach cannot work, as we cannot at any point calculate some partial labeling $\lambda$ of a delta periodic graph $(G,\lambda)$,
    %that could at the end be extended to the whole $G$.

@@

To overcome the above mentioned difficulties, we propose the following high-level strategy.
\begin{itemize}
    \item We identify a small number of ``interesting vertices'' and specific ``graph constructions'' between which we guess representative fastest paths.
    \item For each guess %(also called ``paths configuration'') 
    we create one Integer Linear Program (ILP).
    \item We show that we can determine most labels along each path up to an additive constant. We show that the number of labels we cannot determine is small. 
    \item For each additive constant and each non-determined label we introduce a variable for an Integer Linear Program (ILP).
    \item We use equality and inequality constraints in each ILP to ensure that the fastest temporal path from $u$ to $v$ is of duration $D_{u,v}$, while any other path is of duration at least $D_{u,v}$.
    \item We compute a solution to each of the ILPs (if it exists) and construct a labeling from that solution. Since each ILP has a few variables and not too many constraints, this computation can be done efficiently using Lenstra's Theorem~\cite{Lenstra1983Integer}.
    If none of the ILP instances give a positive solution, then there exists no labeling realizing given matrix.
    This is true because considering all guesses of fastest temporal paths is equivalent to exhaustively searching through all possible temporal paths between vertices.
\end{itemize}

@@

%Let us remember the following basic property of the problem \deltaExact.
Recall that a labeling $\lambda$ of $G$ satisfies $D$ if the duration of a fastest temporal path from vertex $v_i$ to $v_j$ equals $D_{v_i, v_j}$.
%, and no other temporal path from $v_i$ to $v_j$ is faster.
In order to find a labeling that satisfies this property we iterate through all pairs of vertices $v_i, v_j \in V(G)$ and 
first guess the underlying path $P$ of a fastest temporal path from $v_i$ to $v_j$, which results in one equality constraint for the labels on the path $P$. 
For all other temporal paths from $v_i$ to $v_j$ we know that they cannot be faster, so we introduce inequality constraints for them.
This results in producing $poly(n,m)$ constraints. 
It is important to point out that the number of constraint and number of variables are not related, while we introduced all of the necessary constraints, we manage to create only a small number of variables (upper-bounded by some function in $k$).
%Note that we have to do this while keeping the total number of variables upper-bounded by some function in $k$, therefore we do not 
%
For an easier understanding and the analysis of the algorithm we give the following definition.
\begin{definition}\label{def:segments}
	Let $U \subseteq V(G')$ be a set of vertices of interest and let $u,v \in U$.
	A path $P = (u=v_1,v_2, \dots, v_p = v)$  in graph $G'$, where all inner vertices are not in $U$, \ie $v_i \notin U$ for all $i \in \{ 2, 3, \dots, p-1\}$,
	is called a \emph{segment} from $u$ to $v$. We denote it as $S_{u,v}$.
\end{definition}
%From the definition it follows, that there are $O(k^2)$ segments in $G'$. \todo{True but does not follow directly just from the definition}
Note from \cref{def:segments} we get that $S_{u,v} \neq S_{v,u}$. 
Observe that a temporal path in $G'$ between two vertices of interest is either a segment, or consists of a sequence of some segments. Furthermore, since we have at most $4k$ interesting vertices in $G'$, we can deduce the following important result.
\begin{corollary}\label{obs:FPT-k2segments}
	There are at most $O(k^2)$ segments in $G'$.
\end{corollary}
%In the analysis of the algorithm we rely heavily on the fact that there are $O(k^2)$ segments in $G'$.
%, which we prove later on (see~\cref{obs:FPT-k2segments}).
%For a pair $u,v$ of vertices in $U$, there are $O(k!)=O(k^k)$ possible paths in $G'$ between them.

@@

	Let $S_{u,v} = (u=v_1,v_2, \dots, v_p = v)$ and $S_{w,z} = (w=z_1,z_2, \dots, z_r = z)$ be two segments. %,
	%first one between vertices $u,v \in U$, second one between vertices $w, z \in U$.

@@

	Recall that $S_{u,v}\neq S_{v,u}$ for every $u,v\in U$. Furthermore note that we did not assume that $\{u,v\} \cap \{w,z\} = \emptyset$. Therefore, by repeatedly making the above guesses, we also guess the following fastest temporal paths: 

@@

	and let $w \in U \cup Z^*$. %be either a vertex of interest or a representative vertex of a tree, whose clipped vertex is a vertex of interest.

@@

	For fixed $S_{u,v}$ and $w \in U \cup Z^*$ we have $O(k^k)$ different possible such paths, therefore we make $O(k^{k^4})$ guesses for these paths.

@@

	In the case when $S_{u,v}$ is of length $4$, the fixed vertex $v_i$ is the middle vertex, else we fix an arbitrary vertex $v_i \in S_{u,v} \setminus \{u,v\}$.

@@

	We want to determine the fastest paths from $v_i$ to all inner vertices of $S_{w,z}$. We do this by inspecting the values in matrix $D$ from $v_i$ to inner vertices of $S_{w,z}$.

@@

		In this case we do not introduce any new guesses, as we have aready guessed 

@@

	Let $w \in U \cup Z^*$ 
	and let $S_{u,v} = (u=v_1,v_2, \dots, v_p = v)$.

@@

	All together we make two guesses for each pair of vertex $w \in U$ and segment $S_{u,v}$.
	We repeat this for all vertices of interest, and all segments,
	which produces $O(k^{k^2})$ new guesses.
	For an example see~\cref{fig:FPT-guessG7}.
	Detailed analysis follows arguing from above (as in G-\ref{FPT:guess-splitFromAnotherSegmentAndPaths}) and is deferred to Appendix.
	% counter for guesses continues
\begin{figure}[t]
	\centering
	\begin{subfigure}[b]{0.48\textwidth}
		\centering
		\resizebox{\linewidth}{!}{
			\includegraphics{fig-guess1}}
		\caption{Example of an Inner segment guess I (G-\ref{FPT-guessFTPamongv2z2}), where we guessed the fastest temporal paths of the form $v_2 \rightarrow u \leadsto w \rightarrow z_2$ (in blue)
			and $v_2 \rightarrow u \leadsto z \rightarrow z_{r-1}$ (in red).
			\label{fig:FPT-guessG4}}
	\end{subfigure}
	\quad
	\begin{subfigure}[b]{0.48\textwidth}
		\centering
		\resizebox{\linewidth}{!}{

			\includegraphics{fig-guess2}}
		\caption{Example of an Inner segment guess II (G-\ref{FPT:guess-uToSegmentz2}), where we guessed the fastest temporal paths of the form $w \leadsto u \rightarrow v_2$ (in blue) and $w \leadsto v \rightarrow v_{p-1}$ (in red). 
			\label{fig:FPT-guessG5}}
	\end{subfigure}

	\begin{subfigure}[b]{0.48\textwidth}
		\centering
		\resizebox{1.2\linewidth}{!}{
			%
			\includegraphics{fig-guess3}}
		\caption{Example of a Split vertex guess I (G-\ref{FPT:guess-splitFromAnotherSegmentAndPaths}), where, for a fixed vertex $v_i \in S_{u,v}$,
			we calculated its corresponding split vertex $z_j \in S_{w,z}$,
			and guessed the fastest paths of the form
			$v_i \rightarrow v_{i-1} \rightarrow \cdots \rightarrow u \leadsto z \rightarrow z_{r-1} \cdots \rightarrow z_j$ (in blue) 
			and $v_i \rightarrow v_{i+1} \rightarrow \cdots \rightarrow v \leadsto w \rightarrow z_2 \rightarrow \cdots \rightarrow z_{j-1}$ (in red). 
			\label{fig:FPT-guessG6}}
	\end{subfigure}
	\quad
	\begin{subfigure}[b]{0.48\textwidth}
		\centering
		\resizebox{\linewidth}{!}{
			\includegraphics{fig-guess4}}
		\caption{Example of a Split vertex guess II (G-\ref{FPT:guess-splitFromUtoAnotherSegment}), where, for a vertex of interest $w$, 
			we
			calculated its corresponding split vertex $v_i \in S_{u,v}$,
			and guessed the fastest paths of the form
			$w \leadsto u \rightarrow v_2 \rightarrow \cdots \rightarrow v_i$  (in blue) 
			and $w \leadsto v \rightarrow v_{p-1} \rightarrow \cdots \rightarrow v_{i+1}$ (in red). 
			\label{fig:FPT-guessG7}}
	\end{subfigure}
	\caption{An example of guesses G-\ref{FPT-guessFTPamongv2z2}, G-\ref{FPT:guess-uToSegmentz2}, G-\ref{FPT:guess-splitFromAnotherSegmentAndPaths} and G-\ref{FPT:guess-splitFromUtoAnotherSegment}.}
\end{figure}
%
We create all of the above guesses independently.
We start with the first one, that results in $O(k^{k^3})$ different possibilities, then we split each one of these guesses into $O(k^{k^3})$ new ones, that respond to the guessing in the second step, etc.
After creating all of the guesses we end up with $f(k)$ different cases (where $f$ is a double exponential) 
and create one ILP instance for each case.
